---
title: "The Most At-Risk Populations for Chronic Absenteeism in NYC Public Schools"
author: "Anaan Choudhury, Karen Galvan and Naylynn Tañón Reyes"
format:
  html:
    self-contained: true
#bibliography: references.bib

---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r, echo = FALSE}
library(tidyverse)
library(dplyr)
library(Stat2Data)
library(performance)
library(GGally)
library(car)
```

# Abstract

In New York City, chronic absenteeism is defined as missing 18 or more days of school. With students at risk for academic underperformance, especially due to the tumultuous learning environment due to the pandemic, creating equitable education for students is vital. But which populations are most at risk? To answer this question, we compared factors of students’ identities to search for a correlation between certain populations and chronic absenteeism. Our study shows there is a significant correlation between the number of students living in poverty and the number of students who are chronically absent. Additionally, there is a positive correlation between students who are English language learners, living in poverty, and living with disabilities and chronic absenteeism in schools which we explored through an additive model. 



# Introduction 

Over the long term, chronic absenteeism is correlated to increased rates of high school dropout, adverse health outcomes and poverty in adulthood, and an increased likelihood of interacting with the criminal justice system (Center for Research in Education and Policy Research (2018)). In New York City, chronic absenteeism is defined as missing 18 or more days of school. As reported in April of the 2021-22 school year, city data revealed that 37% of New York City students were chronically absent (Zimmerman 2022). Because missing class instruction severely impacts a student's academic performance, we hope to investigate the factors of identity that impact chronic absenteeism in students to better predict which populations are most at risk. By identifying factors that contribute the most to absenteeism, public schools can begin to identify resolutions and more effectively allocate funding and resources to support students. For example, this research can help schools apply for funding for specialized counselors who aid students in getting back into school if they've been absent for long periods of time. They can  stay in contact with students and their parents. Then upon their return these counselors can help them map out the work that needs to be done, due dates and identify what they should prioritize in order to help students not be overwhelmed by the workload after their return. This is especially important for students below the poverty line as, studies show that, education greatly supports social mobility (Center for Research in Education and Policy Research (2018).

  By using the NYC public school system as a sample for the United States public school system, we will answer the question “What variables drive chronic absenteeism in NYC public schools?”. We will be investigating whether certain independent variables impact higher rates of chronic absenteeism. Our data sets include demographic information like ethnicity, gender, non-native English speakers, poverty levels, and disability status at the individual school level.

  We hypothesize that there is a positive correlation between poverty level, disability status, non-native English speaking students and chronic absenteeism rates for students in the NYC public school system. In general, we also suppose that all marginalized students have higher rates of absenteeism in the NYC public school system due to the lack of planning for equity.

# Methods

  We used the data sets provided by NYC Open Data, which is free public data published by New York City agencies and other partners. Specifically, we used the “2016-17 - 2020-21 School End-of-Year Attendance and Chronic Absenteeism Data” and the “NYC Open Data - 2017-18 - 2021-22 Demographic Snapshot”, which we joined by “DBN”, or the district borough number, which uniquely identifies a school in the five borough districts. The former tells us about the total number of chronically absent students for a given school or “DBN”, along with the total number of absences for the year, and the total number and percentage of chronically absent students. The Demographic Snapshot data, on the other hand, provides information on students at the school-wide level, along with representative information pertaining to grade level, sex, ethnicity/race, disability status, English language learners, and poverty level. All investigations were conducted through the Board of Education on the school level, rather than on individual students. 


```{r, echo = FALSE}
#Uploading dataframes and combining them by the "DBN" variable

data <- read.csv("absenteeismData.csv") %>% 
  filter(Grade == "All Grades" & Year =='2020-21')

data2 <- read.csv("demographicData.csv") %>% 
  filter(Year =='2020-21')

```

  After joining both data frames we had to select our variables. The variables used in our analysis include percentages of chronically absent students, poverty level, disability status, and students who are English language learners. All of these are expressed as percentages of students out of the total student bodies at individual New York City public schools (including grades Kindergarten through 12th grade). We specifically chose these variables because of the repeated institutional discrimination against these marginalized groups in the United States throughout history.

  Next, we removed all rows with N/As from the new data frame to clean the data. Then we had to do some formatting changes which included removing percentage signs from the column values. The Poverty and Chronically Absent columns then had their variable type converted from “char” to “num” in order to do calculations and were converted to the same written percentage format for consistency. 

```{r, echo = FALSE}

#remove any duplicate columns
joined_data <- full_join(data, 
                         data2, 
                         by="DBN", 
                         suffix=c("",".y")) %>%
  dplyr::select(-ends_with(".y"))

# remove all rows that have any N/A values from the new data frames
joined_data <- na.omit(joined_data)

# =====Modifying joined dataset======

# rename variables of the columns we are interested in for our models
joined_data <- joined_data %>% 
  rename("Chronically.Absent" = "X..Chronically.Absent.1",
         "Poverty" = "X..Poverty.1",
         "English.Language.Learners" = "X..English.Language.Learners.1",
          "Students.with.Disabilities" = "X..Students.with.Disabilities.1")

# rename every row value that says 'Above 95%' <- '95'
joined_data$Poverty[joined_data$Poverty == 'Above 95%'] <- '95'

# replace the % sign from every row so we are left with just a number
joined_data$Poverty <- gsub("%", "", joined_data$Poverty)

# change the data type from char to num for two columns
joined_data$Poverty = as.numeric(as.character(joined_data$Poverty))
joined_data$Chronically.Absent = as.numeric(as.character(joined_data$Chronically.Absent))

# change 'Chronically.Absent' and 'Poverty' to the same percentage format (0.50 vs 50)
# divide columns by 100 and replace values
joined_data <- joined_data %>% 
   mutate(Poverty = Poverty / 100) %>% 
   mutate(Chronically.Absent = Chronically.Absent / 100)

```

  To begin our data exploration, we created a scatterplot with a regression line of the percentage of chronically absent students against the percentage of students below the poverty line to search for a relationship between the two variables. 

```{r, echo = FALSE}

#Initial data exploration

ggplot(data = joined_data, mapping = aes(x = Poverty, y = Chronically.Absent)) +
  geom_point() +
  geom_smooth(method = lm, se=FALSE, formula = y~x) +
  labs(title = "Students Living in Poverty vs Students Chronically Absent", 
       subtitle = "In NYC schools", 
       x ="Poverty (%)", 
       y ="Chronically Absent (%)")
```

Next, we reported the regression table to interpret the coefficients and p-value of the regression line. Since we saw a p-value smaller than 0.05 and a positive linear correlation between our initial variables, Poverty and Chronically.Absent we decided to continue with this work and add more variables.

```{r, echo = FALSE}
absent_model_1var <- lm(Chronically.Absent ~ Poverty, data = joined_data)
summary(absent_model_1var)$coefficients
```

Because we want to explore the relationships between a school’s reported chronic absenteeism and Poverty, English.Language.Learners and Students.with.Disabilities, a new regression table was created with the additional explanatory variables to check the p-values and remove those with a p-value greater than 0.05. 

```{r, echo = FALSE}
# Adding more coefficients

absent_model_3var <- lm(Chronically.Absent ~ Poverty + 
                                        English.Language.Learners + 
                                        Students.with.Disabilities, 
                                        data = joined_data)
summary(absent_model_3var)$coefficients
```

Since all of the variables have p-values less than 0.05, we can keep all of them for now.

Then a check for multicollinearity was performed to investigate whether certain variables were more strongly correlated with some combination of the other explanatory variables in the model. This was tested by calculating the variance inflation factor (VIF) Generally, VIF values > 10 are  a sign of danger, while VIF values > 5 are a sign  of caution. Here we see that the VIF values for all variables are below 5 therefore we are not concerned about this degree of multicollinearity.


```{r, echo = FALSE}
vif(absent_model_3var)
```

Afterwards, several models were created by removing variables (reducing the complexity) in order to later compare the models. Then, by using the anova() function in R, we used nested F-tests to compare the goodness of fit between each additive model. 


```{r, echo = FALSE}
absent_model_3var <- lm(Chronically.Absent ~ Poverty + 
                                            English.Language.Learners + 
                                            Students.with.Disabilities, 
                                            data = joined_data)
absent_model_2var <- lm(Chronically.Absent ~ Poverty + 
                                             English.Language.Learners, 
                                             data = joined_data)

anova(absent_model_2var, absent_model_3var)

```

The results indicate we should prefer absent_model_3var as the best balance of goodness of fit and model complexity. We then compared that model to the new models with interactions added using an nested F-test once again.

```{r, echo = FALSE}
absent_model_interact1 <- lm(Chronically.Absent ~ Poverty * 
                                                  English.Language.Learners * 
                                                  Students.with.Disabilities, 
                                                  data = joined_data)

absent_model_interact2 <- lm(Chronically.Absent ~ Poverty * 
                                                  English.Language.Learners + 
                                                  Students.with.Disabilities, 
                           data = joined_data)

absent_model_interact3 <- lm(Chronically.Absent ~ Poverty + 
                                                  English.Language.Learners * 
                                                  Students.with.Disabilities, 
                            data = joined_data)
```


```{r, echo = FALSE}
anova(absent_model_3var, absent_model_interact3)
anova(absent_model_3var, absent_model_interact2)
anova(absent_model_3var, absent_model_interact1)

```

 
Here absent_model_interact3 and absent_model_interact1 show evidence that there is an interaction due to the p-values being less than 0.05 but the difference in RSS values is extremely small. This means that even though there is a statistically significant change when applying interaction to the model, due to the low  RSS value, there is no practical change to the estimated rate of absenteeism. This leads us to believe that the interactions are not a necessary component of the model and that absent_model_3var, which is the reduced model, is still the best balance of goodness of fit and model complexity. 


Next we checked to see if our model meets the lineariy, normality and equal variance assumptions. Our model, however, did not pass the tests for normality and equal variance. 
  
```{r, echo = FALSE}
# linearity
absent_model_3var_linearity_check <- check_model(absent_model_3var, 
                                                 check="linearity",
                                                 panel=FALSE
                                                 )
```


```{r, echo = FALSE}
# normality
absent_model_3var_normality_check <- check_model(absent_model_3var, 
                                                 check="qq",
                                                 panel=FALSE
                                                  )
```

```{r, echo = FALSE}
# homoegeneity / normality
absent_model_3var_homoegeneity_check <- check_model(absent_model_3var, 
                                                    check="homogeneity",
                                                    panel=FALSE
                                                    )
```
  

In order to stabilize the variance we decided to apply a log transformation to the outcome variable. However, because several schools had zero percent absenteeism, and zero cannot be converted to the logarithmic scale, we could not apply the log transformation. Rather than omit valid observations and potentially bias our coefficients we chose to fit and interpret the model on the original scale regardless of the normality and equal variance violations. Therefore there is that to consider when reviewing our results. Still, note that T tests on the coefficients of regression models are generally robust to the violations of normality and equal variance. Therefore, it would take an incredibly large violation of normality to get the sampling distribution and give it a different shape so a violation like ours would not dramatically change the shape of the T-distribution. Hence, our model is still valuable and the information gathered from it is still valid.


```{r, echo = FALSE}
#log_absent_model_3var <- lm(log(Chronically.Absent) ~ Poverty + English.Language.Learners + Students.with.Disabilities, data = joined_data)
```


```{r, echo = FALSE}
# linearity
# 
# log_absent_model_3var_linearity_check <- check_model(log_absent_model_3var, check="linearity",
#                               panel=FALSE
#                               )
# plot(log_absent_model_3var_linearity_check, data=joined_data)
```


```{r, echo = FALSE}
# normality
# log_absent_model_3var_normality_check <- check_model(log_absent_model_3var, check="qq",
#                               panel=FALSE
#                               )
# plot(log_absent_model_3var_normality_check, data=joined_data)
```

```{r, echo = FALSE}
# homoegeneity / normality
# log_absent_model_3var_homoegeneity_check <- check_model(log_absent_model_3var, check="homogeneity",
#                               panel=FALSE
#                               )
# plot(log_absent_model_3var_homoegeneity_check, data=joined_data)
```

# Results 

We concluded that there is a positive linear correlation between the outcome variable and all explanatory variables and even though the overall change in the percentage of absenteeism depends on all three variables the effect of each individual variable does not depend on the other. Meaning these three variables, individually, are correlated to higher average rates of absenteeism in NYC public schools. Therefore as each of these variables increase the rate of absenteeism at a school also tends to rise.


###Figure 1: Scatterplot of Students Living in Poverty vs Students Chronically Absent

````{r}
#Figure 1

ggplot(data = joined_data, mapping = aes(x = Poverty, y = Chronically.Absent)) +
  geom_point() +
  geom_smooth(method = lm, se=FALSE, formula = y~x) +
  labs(title = "Students Living in Poverty vs Students Chronically Absent", 
       subtitle = "In NYC schools", 
       x ="Poverty (%)", 
       y ="Chronically Absent (%)")
```

Figure 1 above depicts a clear positive, linear relationship between our variables: the percentage of chronically absent students of the total student body at a particular school, the percentage of students living in poverty of the total student body at a particular school. 

```{r}
absent_model <- lm(Chronically.Absent ~ Poverty, data = joined_data)
summary(absent_model)$coefficients
```


These coefficients represent the unique effect of one explanatory variable on the outcome,
after taking into account the effect of the other explanatory variables

**Intercept:** The predicted reported percentage of 0 chronically absent student population for a school with 0 percent reported poverty rate is -0.0471.

**Poverty:** For each additional 1 unit percent of students under the poverty line, the predicted reported percentage of chronically absent student population for a school increases by 0.476 on average.


By adding more coefficients to this model, we are able to see that there is a clear positive, linear relationship between the variables of students with disabilities, English language learners, and students living in poverty in an additive model.

```{r}
# Adding more coefficients

absent_model2 <- lm(Chronically.Absent ~ Poverty + 
                                        English.Language.Learners + 
                                        Students.with.Disabilities, 
                                        data = joined_data)
summary(absent_model2)$coefficients
```

These coefficients represent the unique effect on one explanatory variable on the outcome, after taking into account the effect of other explanatory variables. 

**Intercept:** The predicted reported percentage of 0 chronically absent student population for a school with 0 percent reported poverty rate, 0 percent reported population of student English language learners, and 0 percent reported population of students with disabilities is -0.081.

**Poverty:** For each additional 1 unit percent of students under the poverty line, the predicted reported percentage of chronically absent student population for a school increases by 0.453 on average.

**English Language Learners:** For each 1 percent increase in the student population who are english language learners, the predicted reported percentage of chronically absent student population for a school decreases by -0.101 on average, regardless of simultaneous changes in the percentage of students with disabilities

**Students with Disabilities** For each 1 percent increase in the student population who have disabilities, the predicted reported percentage of chronically absent student population for a school increases by 0.270 on average, regardless of simultaneous changes in the percentage of enlgish language learners.

###Figure 2: Scatterplot Matrix 

Figure 2 is a visualization that maps out the variables in an additive manner. The matrix plots all of our variables against each other in individual plots, so the first column plots 'chronically absent' vs poverty, then 'english language learner' then 'student with disabilities'. It also plots a histogram of each variable and the correlation between variables.


```{r}

#Figure 2: Scatterplot Matrix 


data_minimal <- joined_data %>% 
  select(Chronically.Absent,Poverty, 
         English.Language.Learners, 
         Students.with.Disabilities)

ggpairs(data_minimal)  +
  labs(title = "Scatterplot Matrix of Our Model", 
       subtitle = "", 
       x ="", 
       y ="")
```

 
# Discussion

In this investigation we set out to answer “What variables drive chronic absenteeism in NYC public schools?”. By creating additive and interactive models, we found the best combination of the variables of the percentages of students living in poverty, percentages of English language learners, and the percentage of students with disabilities out of the total student body at individual public schools in New York City to explain chronic absenteeism. We conclude that an additive model accounting for the percentages of students living in poverty, percentages of English language learners, and the percentage of students with disabilities best explains chronic absenteeism because it has the best fit to explain the trend of these variables in scatterplots tested through Nested F-tests. 
The limitations of this study include the failure of the normality check and equal variance check. In order to stabilize the variance we applied log to the outcome variable. However, because several schools had zero percent absenteeism and zero cannot be converted to the logarithmic scale we could not apply log transformation. It should be noted that we did not convert the scale of our findings to log after contemplating whether the consequences of omitting valid observations and potentially creating a bias in our coefficients since many schools had zero percent absenteeism and zero cannot be converted to the logarithmic scale
The strength of our study lies in the integrity of the New York City Department of Education’s comprehensive data collection on NYC public schools. We checked for patterns in the missing data cells labeled as N/A and found none. We deduce that there is a strong correlation not causation between the explanatory variables and chronic absenteeism in this study. The causation is due to a lack of resources for marginalized students and a lack of funding for public education in New York City.
 

# Data Analysis Appendix

The Fitted vs Residuals plot applies to both the linearity assumption as well as the equal variance/homogeneity assumption. When looking at this plot, we can see that the points do have a cone shape, therefore it does not pass the equal variance test. When looking for linearity we want to see that there is no relationship between the values and the residuals which we see in this plot the points are not randomly spread apart. Therefore, both assumptions are not met.

Seeing the Normality plot, we can see the tail on the right of our plot does not fall along line so there is some skew, still this skew is not large enough for us to consider this assumption failed.

Our model’s failure of the normality and equal variance check are not completely negative outcomes. As stated before, we specifically care about the shape of the sampling distribution of the T statistic. The violation Since it would take an incredibly large violation of normality to get the sampling distribution and give it a different shape, a violation like ours would not dramatically change the shape of the T-distribution. 

###Linearity Check

```{r}

plot(absent_model_3var_linearity_check, data=joined_data)
```

###Normality Check

```{r}
plot(absent_model_3var_normality_check, data=joined_data)

```


###Homogeneity Check
```{r}
plot(absent_model_3var_homoegeneity_check, data=joined_data)
```




# Citations

Center for Research in Education and Policy Research (2018). *Chronic Absenteeism and Its Impact on Achievement. CRESP Policy Brief Series.* Retrieved 2022, from <https://www.cresp.udel.edu/wp-content/uploads/2018/07/P18-002.5_final.pdf> 


Chetty, R., Hendren, N., Jones, M., & Porter, S. (2020). *Race and Economic Opportunity in the United States: An intergenerational perspective.* Race and Economic Opportunity in the United States: An Intergenerational Perspective. Retrieved December 15, 2022, from <https://scholar.harvard.edu/hendren/publications/race-and-economic-opportunity-united-states-intergenerational-perspective> 

Nauer, K., Mader, N., Robinson, G., Jacobs, T., Cory, B., Moss, J., & Bloodworth, A. (2014). *A Better Picture of Poverty What Chronic Absenteeism and Risk Load Reveal About NYC's Lowest-Income Elementary Schools.* A Better Picture of Poverty | Attendance Works. Retrieved 2022, from <http://static1.squarespace.com/static/53ee4f0be4b015b9c3690d84/t/547ccfefe4b03f842b275001/1417465839022/Better+Picture+of+Poverty_PA_FINAL.pdf> 

Zimmerman, A. (2022, September). *41% Of NYC students were chronically absent last school year.* Chalkbeat New York. <https://ny.chalkbeat.org/2022/9/16/23357144/chronic-absenteeism-pandemic-nyc-school#:~:text=About%2041%25%20of%20New%20York,to%20city%20data%20released%20Friday.> 